Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics in Australia

### 2024-11-06 paper/paper-anzsj.tex
shifted tables and made fig 1 into a single image

\begin{document}
\maketitle
\begin{abstract}
The choropleth map display is commonly used for communicating spatial
distributions across geographic areas. However, when choropleths are
used the size of the geographic units will influence the understanding
of the distribution derived by map users. The hexagon tile map is
presented as an alternative display for effectively visualizing population related
distributions. Visual inference is used to measure the power
of the hexagon tile map design, and the choropleth is used as a
comparison. The hexagon tile map display is tested using a distribution
that is directly related to the geography, with values monotonically
increasing from the North-West to South-East areas of Australia. This
study finds in a hexagon tile map lineup the single map that contains a
population-related distribution is detected with greater probability
than the same data displayed in a choropleth map. These findings should
encourage map creators to implement alternative displays and consider a
hexagon tile map when presenting spatial distributions of heterogeneous
areas.
\end{abstract}

@@

{\centering 
\includegraphics[width=0.46\linewidth]{paper_files/figure-latex/thyroid-choro-1} 
\includegraphics[width=0.54\linewidth]{paper_files/figure-latex/thyroid-hex-1} 
}

@@

The choropleth map is an effective spatial display if the size of the
geographic units is relatively uniform. This is not the case for most
countries. Size heterogeneity in administrative units is particularly
extreme in Australia: most of the landscape of Australia is sparsely
settled, with the population densely clustered into the narrow coastal
strips. Fig. \ref{fig:thyroid} shows the choropleth map of thyroid
cancer rates in Australia. The choropleth map focuses attention on the
geography, and for heterogeneously sized areas it presents a biased view
of the population-related distribution of the statistic
\citep{CBATCC}. \emph{Land doesn't get cancer, people
do} -- a more effective way to communicate the spatial distributions of
cancer statistics is needed.

@@

The hexagon tile map algorithm, automatically matches spatial regions to
their nearest hexagon tile, from a grid of tiles. It has the effect of
spreading out the inner city areas while maintaining the spatial
locations or regions in remote areas. The algorithm is available in the
R package,  \texttt{sugarbag} \citep{sugarbag}. Fig.
\ref{fig:thyroid} shows the hexagon tile map. Colour maps from
substantially below average (blue) to substantially above average (red)
rates. The inner city areas have expanded, making it possible to see the
cancer incidence in the small, densely populated areas. Remote regions
are represented by isolated hexagons, which is not ideal, but maintains
the spatial location of these data values. It is of interest to know how
well the spatial distribution is perceived for this display, in
comparison to the choropleth.

@@

The chance that an observer picks the data plot out of a lineup of size
\(m\) plots accidentally, if the null hypothesis is true is \(1/m\).
With \(K\) observers, the probability of \(k\) randomly choosing the
data plot, roughly follows a binomial distribution with \(p=1/m\). Fig.
\ref{fig:lineup} shows a lineup of the hexagon tile map, of size
\(m=12\). Plot 3 is the data plot, and the remaining 11 are plots of
null data.

@@

The primary factor in the experiment is the plot type. The secondary
factor is a trend model. Three trend models were developed, one
mirroring a large spatial trend for which the choropleth would be
expected to do well, and two with differing levels of inner-city hot
spots. These latter two reflect the structure seen in thyroid cancer
data (Fig. \ref{fig:thyroid}). This
produces six treatment levels:

@@

Data is generated for each of the trend models, with four replicates,
and each displayed both as a choropleth and as a hexagon tile map, which
yields 12 data sets, and 24 data plots. This set of displays is divided
in half, providing two sets of 12 displays, Group A and Group B.
Participants were randomly allocated to Group A or B. Participants saw a
data set only once, either as a choropleth or as a hexagon tile map.
Table \ref{fig:exp-design} summarises the design and the allocation of
the displays.

@@

This smoothing allowed neighbors to be related to each other, but also
allowed outliers, and showed distributions similar to the thyroid cancer
distribution (Fig. \ref{fig:thyroid}).

@@

Fig. \ref{fig:detect-compare} displays the average detection rates for
the two types of plot separately for each trend model. Each trend model
was tested using four repetitions, evaluations on the same data set were
seen as either choropleths or hexagon tile maps by each group as
specified in Table. \ref{fig:exp-design}; the detection rates for each
display are connected by a line segment. The Three Cities and All Cities
trend models shown in the hexagon tile map allowed viewers to detect the
data plot substantially more often than the choropleth counterparts. One
replicate for the All Cities group had similar detection rates for both
plot types, the rate of detection using the choropleth map was much
higher than other replicates. Surprisingly, participants could also
detect the gradual spatial trend in the NW-SE group from the hexagon
tile map. We expected that the choropleth map would be superior for the
type of spatial pattern, but the data suggests the hexagon tile map
performs slightly better, or equally as well.

@@

Table. \ref{tab:desc-stats} shows the means and standard deviations of
the detection rate for each type of plot and each trend model. This also
gives the standard deviations, the smallest standard deviation for all
sets of replicates was the Three Cities trend model shown in a
Choropleth display. This group of displays had a very small detection
rate of 0.04. The mean detection rate for the Three Cities trend model
shown as choropleth map lineups was also the smallest at 0.40. The
North-West to South-East (NW-SE) trend model unexpectedly had a higher
mean detection rate for the hexagon tile map displays, but the
difference in the means of detection rate was only 0.10.

\begin{table}

Table. \ref{tab:detect-glmer1} presents a summary of the generalised
linear mixed effects model, testing the effect of plot type and trend
model on the detection rate. The results support the summary from Fig.
\ref{fig:detect-compare} and all parameters are statistically
significant despite the large standard deviations observed in Table.
\ref{tab:desc-stats}. Overall, the hexagon tile map performs marginally
better than the choropleth for all trend models, which is a pleasant
surprise. Allowing for the interaction effect, the difference in
detection rate decreases for population-related displays for a
choropleth map lineup but increases for a hexagon tile map display. The
log odds of detection shown in Table. \ref{tab:detect-glmer1} can be back-transformed after taking the sum of all terms for the trend and type of
display that are of interest. For the NW-SE distribution, the predicted
detection rate for the hexagon tile map display increases the predicted
probability of detection to 0.63 from 0.52 for choropleths, this is
almost exactly the difference seen in the table of means and is
significant only at the 0.05 level.

@@

\begin{table}

@@

Fig. \ref{fig:beeswarm} shows horizontally jittered dot plots to
contrast the time taken by participants to evaluate each lineup when
viewing each type of display. The time is also separated by the trend model
and whether the data plot was detected or not detected. The time taken
to complete an evaluation ranged from milliseconds to 60 seconds. The
average time taken for each type of display is shown as a large coloured dot
on each plot. when considering the heights of the green and orange dots,
there is little difference in the average time taken to read a
choropleth or hexagon tile map. Comparing the same coloured dot across
each trend model row, there is a slight increase in the time taken to
correctly detect the data plot in the hexagon tile map lineup, but
little difference in evaluation time for the choropleth display.
However, there were substantially fewer correct detections for choropleth
lineups for the Three cities and All Cities trends.

@@

Participants provided their level of certainty regarding their choice
using a five-point scale. Unlike the accuracy and speed of responses
that were derived during the data processing phase, this was a
subjective assessment by the participant prompted by the question: `How
certain are you about your choice?'. Fig. \ref{fig:certainty} shows the
number of times participants provided each level of certainty. This was
separated for each combination of trend models and display type, and
colored depending on whether a participant correctly detected the data
plot in the lineup. Participants often chose 4 or 5 when viewing the
population-related trends in the choropleth display, even though they
were often incorrect when viewing an All Cities trend and overwhelmingly
incorrect for the Three Cities trend. This shows overconfidence in their
detection ability when using a choropleth map display. Participants were
less likely to be certain when their choice was incorrect and they were
viewing a hexagon tile map. For each trend model, participants were more
likely to doubt their choice and choose 1 or 2 in the hexagon tile map
displays, even though many had made the correct choice.

@@

We expected that the choropleth map would be superior for communicating
the spatial pattern of geographic distributions. The data suggest that
the participants perform slightly better or equally as well for each
replicate in each trend model across the two displays. Table II shows
that the difference in the mean detection rate for the two trend models
was 0.10.

The differences are seen in the \ref{fig:detect-compare} plot and Table.
\ref{tab:desc-stats} are reflected in the model results. Surprisingly
the difference in the geographic distribution was significant at the
0.05 level. It also showed that the hexagon tile map display performs
marginally better than the choropleth for all trend models. Unexpectedly
the detection rate suffers when using a choropleth map to display
population-related distributions.

While the significance of the difference in detection was the key focus
of this experiment, the secondary focus was the time taken by
participants. it was expected that the participants might take longer to
consider the hexagon tile map distribution but would be able to detect
the data plot in the lineup. The bimodal distributions seen in Fig.
\ref{fig:beeswarm} showed very little difference in the mean evaluation
times. As the maximum time of all of the distributions approached 60
seconds it cannot be said that the participants took longer to evaluate
the hexagon tile map displays.

The responses to the questions asked of participants included the reason
for their choice and the certainty around their choice. Fig.
\ref{tab:reason} shows high levels of certainty of 4 and 5 were chosen
by participants when looking at the population distributions in a
choropleth map display showing that they were overconfident when
attempting to find the real data plot in the choropleth map displays.
Participants performed better on the NW-SE distribution shown in the
choropleth display and were reasonably confident about their decisions.
The high levels of the mid-range value of 3 could indicate that the
participant did not want to provide a response, as this was the default
value. Those who chose level 4 or 5 were equally likely to be correct
for the three cities lineups, but more likely to be correct than
incorrect for the other two trend models.

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
The choropleth map display is commonly used for communicating spatial
distributions across geographic areas. However, when choropleths are
used the size of the geographic units will influence the understanding
of the distribution derived by map users. The hexagon tile map is
presented as an alternative display for effectively visualizing population-related
distributions. Visual inference is used to measure the power
of the hexagon tile map design, and the choropleth is used as a
comparison. The hexagon tile map display is tested using a distribution
that is directly related to the geography, with values monotonically
increasing from the North-West to South-East areas of Australia. This
study finds in a hexagon tile map lineup the single map that contains a
population-related distribution is detected with greater probability
than the same data displayed in a choropleth map. These findings should
encourage map creators to implement alternative displays and consider a
hexagon tile map when presenting spatial distributions of heterogeneous
areas.
\end{abstract}

@@

{\centering 
%\includegraphics[width=0.46\linewidth]{paper_files/figure-latex/thyroid-choro-1} 
%\includegraphics[width=0.54\linewidth]{paper_files/figure-latex/thyroid-hex-1} 
\includegraphics[width=\linewidth]{figures/thyroid.pdf}
}

@@

The choropleth map is an effective spatial display if the size of the
geographic units is relatively uniform. This is not the case for most
countries. Size heterogeneity in administrative units is particularly
extreme in Australia: most of the landscape of Australia is sparsely
settled, with the population densely clustered into the narrow coastal
strips. Figure \ref{fig:thyroid} shows the choropleth map of thyroid
cancer rates in Australia. The choropleth map focuses attention on the
geography, and for heterogeneously sized areas it presents a biased view
of the population-related distribution of the statistic
\citep{CBATCC}. \emph{Land doesn't get cancer, people
do} -- a more effective way to communicate the spatial distributions of
cancer statistics is needed.

@@

The hexagon tile map algorithm, automatically matches spatial regions to
their nearest hexagon tile, from a grid of tiles. It has the effect of
spreading out the inner city areas while maintaining the spatial
locations or regions in remote areas. The algorithm is available in the
R package,  \texttt{sugarbag} \citep{sugarbag}. Figure
\ref{fig:thyroid} shows the hexagon tile map. Colour maps from
substantially below average (blue) to substantially above average (red)
rates. The inner city areas have expanded, making it possible to see the
cancer incidence in the small, densely populated areas. Remote regions
are represented by isolated hexagons, which is not ideal, but maintains
the spatial location of these data values. It is of interest to know how
well the spatial distribution is perceived for this display, in
comparison to the choropleth.

@@

The chance that an observer picks the data plot out of a lineup of size
\(m\) plots accidentally, if the null hypothesis is true is \(1/m\).
With \(K\) observers, the probability of \(k\) randomly choosing the
data plot, roughly follows a binomial distribution with \(p=1/m\). Figure
\ref{fig:lineup} shows a lineup of the hexagon tile map, of size
\(m=12\). Plot 3 is the data plot, and the remaining 11 are plots of
null data.

@@

The primary factor in the experiment is the plot type. The secondary
factor is a trend model. Three trend models were developed, one
mirroring a large spatial trend for which the choropleth would be
expected to do well, and two with differing levels of inner-city hot
spots. These latter two reflect the structure seen in thyroid cancer
data (Figure \ref{fig:thyroid}). This
produces six treatment levels:

@@

Data is generated for each of the trend models, with four replicates,
and each displayed both as a choropleth and as a hexagon tile map, which
yields 12 data sets, and 24 data plots. This set of displays is divided
in half, providing two sets of 12 displays, Group A and Group B.
Participants were randomly allocated to Group A or B. Participants saw a
data set only once, either as a choropleth or as a hexagon tile map.
Figure \ref{fig:exp-design} summarises the design and the allocation of
the displays.

@@

This smoothing allowed neighbors to be related to each other, but also
allowed outliers, and showed distributions similar to the thyroid cancer
distribution (Figure \ref{fig:thyroid}).

@@

Figure \ref{fig:detect-compare} displays the average detection rates for
the two types of plot separately for each trend model. Each trend model
was tested using four repetitions, evaluations on the same data set were
seen as either choropleths or hexagon tile maps by each group as
specified in Figure \ref{fig:exp-design}; the detection rates for each
display are connected by a line segment. The Three Cities and All Cities
trend models shown in the hexagon tile map allowed viewers to detect the
data plot substantially more often than the choropleth counterparts. One
replicate for the All Cities group had similar detection rates for both
plot types, the rate of detection using the choropleth map was much
higher than other replicates. Surprisingly, participants could also
detect the gradual spatial trend in the NW-SE group from the hexagon
tile map. We expected that the choropleth map would be superior for the
type of spatial pattern, but the data suggests the hexagon tile map
performs slightly better, or equally as well.

@@

Table \ref{tab:desc-stats} shows the means and standard deviations of
the detection rate for each type of plot and each trend model. This also
gives the standard deviations, the smallest standard deviation for all
sets of replicates was the Three Cities trend model shown in a
Choropleth display. This group of displays had a very small detection
rate of 0.04. The mean detection rate for the Three Cities trend model
shown as choropleth map lineups was also the smallest at 0.40. The
North-West to South-East (NW-SE) trend model unexpectedly had a higher
mean detection rate for the hexagon tile map displays, but the
difference in the means of detection rate was only 0.10.

Table \ref{tab:detect-glmer1} presents a summary of the generalised
linear mixed effects model, testing the effect of plot type and trend
model on the detection rate. The results support the summary from Figure
\ref{fig:detect-compare} and all parameters are statistically
significant despite the large standard deviations observed in Table
\ref{tab:desc-stats}. Overall, the hexagon tile map performs marginally
better than the choropleth for all trend models, which is a pleasant
surprise. Allowing for the interaction effect, the difference in
detection rate decreases for population-related displays for a
choropleth map lineup but increases for a hexagon tile map display. The
log odds of detection shown in Table \ref{tab:detect-glmer1} can be back-transformed after taking the sum of all terms for the trend and type of
display that are of interest. For the NW-SE distribution, the predicted
detection rate for the hexagon tile map display increases the predicted
probability of detection to 0.63 from 0.52 for choropleths, this is
almost exactly the difference seen in the table of means and is
significant only at the 0.05 level.

@@

@@

Figure \ref{fig:beeswarm} shows horizontally jittered dot plots to
contrast the time taken by participants to evaluate each lineup when
viewing each type of display. The time is also separated by the trend model
and whether the data plot was detected or not detected. The time taken
to complete an evaluation ranged from milliseconds to 60 seconds. The
average time taken for each type of display is shown as a large coloured dot
on each plot. when considering the heights of the green and orange dots,
there is little difference in the average time taken to read a
choropleth or hexagon tile map. Comparing the same coloured dot across
each trend model row, there is a slight increase in the time taken to
correctly detect the data plot in the hexagon tile map lineup, but
little difference in evaluation time for the choropleth display.
However, there were substantially fewer correct detections for choropleth
lineups for the Three cities and All Cities trends.

@@

Participants provided their level of certainty regarding their choice
using a five-point scale. Unlike the accuracy and speed of responses
that were derived during the data processing phase, this was a
subjective assessment by the participant prompted by the question: `How
certain are you about your choice?'. Figure \ref{fig:certainty} shows the
number of times participants provided each level of certainty. This was
separated for each combination of trend models and display type, and
colored depending on whether a participant correctly detected the data
plot in the lineup. Participants often chose 4 or 5 when viewing the
population-related trends in the choropleth display, even though they
were often incorrect when viewing an All Cities trend and overwhelmingly
incorrect for the Three Cities trend. This shows overconfidence in their
detection ability when using a choropleth map display. Participants were
less likely to be certain when their choice was incorrect and they were
viewing a hexagon tile map. For each trend model, participants were more
likely to doubt their choice and choose 1 or 2 in the hexagon tile map
displays, even though many had made the correct choice.

@@

We expected that the choropleth map would be superior for communicating
the spatial pattern of geographic distributions. The data suggest that
the participants perform slightly better or equally as well for each
replicate in each trend model across the two displays. Table \ref{tab:desc-stats} shows
that the difference in the mean detection rate for the two trend models
was 0.10.

The differences are seen in Figure \ref{fig:detect-compare} and Table
\ref{tab:desc-stats} are reflected in the model results. Surprisingly
the difference in the geographic distribution was significant at the
0.05 level. It also showed that the hexagon tile map display performs
marginally better than the choropleth for all trend models. Unexpectedly
the detection rate suffers when using a choropleth map to display
population-related distributions.

While the significance of the difference in detection was the key focus
of this experiment, the secondary focus was the time taken by
participants. it was expected that the participants might take longer to
consider the hexagon tile map distribution but would be able to detect
the data plot in the lineup. The bimodal distributions seen in Figure
\ref{fig:beeswarm} showed very little difference in the mean evaluation
times. As the maximum time of all of the distributions approached 60
seconds it cannot be said that the participants took longer to evaluate
the hexagon tile map displays.

The responses to the questions asked of participants included the reason
for their choice and the certainty around their choice. Figure
\ref{tab:reason} shows high levels of certainty of 4 and 5 were chosen
by participants when looking at the population distributions in a
choropleth map display showing that they were overconfident when
attempting to find the real data plot in the choropleth map displays.
Participants performed better on the NW-SE distribution shown in the
choropleth display and were reasonably confident about their decisions.
The high levels of the mid-range value of 3 could indicate that the
participant did not want to provide a response, as this was the default
value. Those who chose level 4 or 5 were equally likely to be correct
for the three cities lineups, but more likely to be correct than
incorrect for the other two trend models.

@@

\newpage

\begin{table}[t]

\begin{table}[t]

\begin{table}[t]
